{fairmetrics}: An R package for fairness evaluation metrics with confidence intervals

### 2025-05-29 paper.Rmd
add citations

Machine Learning (ML) has been increasingly integrated into high-stakes domains that significantly impact individuals and society, including criminal justice, healthcare, finance, employment, and education. As ML systems become a commonplace in decision-making processes, mounting evidence suggest that they often exhibit bias towards gender and ethnicity. For example, in criminal justice systems, COMPAS, a widely used risk assessment tool in the US, incorrectly classified Black defendants as high-risk at nearly twice the rate of White defendants. As COMPAS was used in court to inform judicial decision-making, this bias impacted Black defendants by potentially leading to harsher bail decisions, longer sentences, and reduced parole opportunities compared to White defendants with similar risk profiles. Similarly, a healthcare algorithm deployed throughout the US to identify patients with complex health needs has shown significantly lower accuracy for Black patients relative to White patients. This caused Black patients with equivalent health conditions to be under-referred for essential care services compared to White patients. Therefore, there is a urgent need for companies, engineers, and researchers to ensure that ML models are fair and equitable across different groups before they are deployed in real-world applications.

@@

The `{fairmetrics}` package is designed to evaluate group fairness in the setting of binary classification with a binary protected attribute. This restriction reflects standard practice in the fairness literature and is motivated by several considerations. First, binary classification remains prevalent in many high-stakes applications, such as loan approval, hiring decisions, and disease screening, where outcomes are typically framed as accept/reject or positive/negative. Second, group fairness is the most widely used framework for binary classification tasks. Third, when protected attributes have more than two categories, there is no clear consensus on how to evaluate group fairness. This focus enables {fairmetrics} to provide statistically grounded uncertainty quantification for group fairness metrics commonly applied in binary classification tasks across diverse application domains.

>>>>

Machine Learning (ML) has been increasingly integrated into high-stakes domains that significantly impact individuals and society, including criminal justice, healthcare, finance, employment, and education[@mehrabi_survey_21]. As ML systems become a commonplace in decision-making processes, mounting evidence suggest that they often exhibit bias towards gender and ethnicity. For example, in criminal justice systems, COMPAS, a widely used risk assessment tool in the US, incorrectly classified Black defendants as high-risk at nearly twice the rate of White defendants[@mattuMachineBias]. As COMPAS was used in court to inform judicial decision-making, this bias impacted Black defendants by potentially leading to harsher bail decisions, longer sentences, and reduced parole opportunities compared to White defendants with similar risk profiles. Similarly, a healthcare algorithm deployed throughout the US to identify patients with complex health needs has shown significantly lower accuracy for Black patients relative to White patients[@obermeyerDissectingRacialBias2019]. This caused Black patients with equivalent health conditions to be under-referred for essential care services compared to White patients. Therefore, there is a urgent need for companies, engineers, and researchers to ensure that ML models are fair and equitable across different groups before they are deployed in real-world applications.

@@

The `{fairmetrics}` package is designed to evaluate group fairness in the setting of binary classification with a binary protected attribute. This restriction reflects standard practice in the fairness literature and is motivated by several considerations. First, binary classification remains prevalent in many high-stakes applications, such as loan approval, hiring decisions, and disease screening, where outcomes are typically framed as accept/reject or positive/negative[@mehrabi_survey_21]. Second, group fairness is the most widely used framework for binary classification tasks[@mehrabi_survey_21]. Third, when protected attributes have more than two categories, there is no clear consensus on how to evaluate group fairness. This focus enables {fairmetrics} to provide statistically grounded uncertainty quantification for group fairness metrics commonly applied in binary classification tasks across diverse application domains.
